The `zoon` R package for reproducible and shareable species distribution modelling

### 2016-11-30 ms.Rmd
Minor edits to text plus add threshold to Feng example

Computational sciences are then presented with a modern version of the reproducibility problem as computational research is largely inaccessible let alone reusable.
If research materials and outputs are shared, they are rarely disseminated in a readily modifiable or extensible format despite the great potential for distributing these resources digitally (e.g. figshare, GitHub).
The `zoon` R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
The module-workflow structure enables scientists to more easily to create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure 1).

@@

Examples of subject specific frameworks are found within the Bioinformatics community, for example the Bioconductor infrastructure that uses R (@Gentleman2004, `bioconductor.org`), and BioJS, a repository for JavaScript components (@Corpas2014, `biojs.net`).
Both software aim to increase the compatibility and sharing of computational tools in order to increase the reuse and robustness of tools.
In terms of robustness, the CHASTE project for cardiac modelling (@Mirams2013, `cs.ox.ac.uk/chaste`) has developed into a model repository that offers community benchmarks for the last 50 years of cardiac modelling [@Cooper2016].
These frameworks increase the access, usability and robustness of ultimately scientific research.
Of course, R users could point to the R environment itself, but the stringent requirements of R packages are likely to prevent much of the SDM community from sharing code in a discoverable or usable format.

@@

Unlike software packages for programming languages however, The Zoon Project enforces a modular structure that ensures interoperability between contributed software modules.
This also helps to lower the bar for creating modules (each module is simply an R function with a set of required inputs and outputs) and facilitates automated testing of modules ot ensure software quality.

@@

Many SDM analyses require running similar workflow in parallel. 
In applied settings this may be running the same analysis for multiple species.
In methodological work, it is important to compare new methods or models to benchmarks.
To divide a `zoon` workflow into multiple parallel analyses the `list` function is used.
`list` can be used to split the analysis at any point: listing multiple occurrence modules together will run the workflow on different occurrence datasets while listing multiple model modules will fit different models to a shared, identical dataset but then apply output modules to each model separately.

The complete workflow, with multiple modules defined with the `Chain` function and parallel analyses being defined with the `list` function can be visualised by calling `plot` with the workflow object as it's argument.
This displays the names of all modules and the broader structure of the workflow as well as displaying which modules are not available in the online repository (Figure ).

@@

@@

`zoon` provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform figshare. `ZoonFigshare` takes a workflow object and some minimal metadata and uploads the workflow as an RData object to the user's figshare profile, for others to download, inspect and modify.

@@

```{r fengworkflow, eval = TRUE, dpi = 600, fig.show = "hide", results = "hide", message = FALSE, fig.height = 9, fig.width = 9}
FengPapes <- 
  workflow(occurrence = SpOcc('Dasypus novemcinctus',
                              extent = c(-130, -20, -60, 60)),
            covariate = Bioclim(extent = c(-130, -20, -60, 60),
                                layers = c(1:4, 6, 9, 10, 12, 15)),
              process = Chain(Clean, 
                              MESSMask,  
                              Background(n = 10000, bias = 200), 
                              Crossvalidate(k = 5)),
                model = MaxEnt,
               output = PrintMap(points = FALSE))
```

>>>>

Computational sciences are then presented with a modern version of the reproducibility problem [reference?] as computational research is largely inaccessible let alone reusable.
If research materials and outputs are shared, they are rarely disseminated in a readily modifiable or extensible format despite the great potential for distributing these resources digitally (e.g. figshare, GitHub).
The `zoon` R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
The module-workflow structure enables scientists to more easily to create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure 1).

@@

Examples of subject specific frameworks are found within the Bioinformatics community, for example the Bioconductor infrastructure that uses R (@Gentleman2004, `bioconductor.org`), and BioJS, a repository for JavaScript components (@Corpas2014, `biojs.net`).
Both software aim to increase the compatibility and sharing of computational tools in order to increase the reuse and robustness of tools.
In terms of robustness, the CHASTE project for cardiac modelling (@Mirams2013, `cs.ox.ac.uk/chaste`) has developed into a model repository that offers community benchmarks for the last 50 years of cardiac modelling [@Cooper2016].
These frameworks increase the access, usability and robustness of scientific research.
Of course, R users could point to the R environment itself, but the stringent requirements of R packages are likely to prevent much of the SDM community from sharing code in a discoverable or usable format.

@@

Unlike software packages for programming languages however, The Zoon Project enforces a modular structure that ensures interoperability between contributed software modules.
This also helps to lower the bar for creating modules (each module is simply an R function with a set of required inputs and outputs) and facilitates automated testing of modules to ensure software quality.

@@

Many SDM analyses require running similar workflow in parallel. 
In applied settings this may be running the same analysis for multiple species.
In methodological work, it is important to compare new methods or models to benchmarks.
To divide a `zoon` workflow into multiple parallel analyses the `list` function is used.
`list` can be used to split the analysis at any point: listing multiple occurrence modules together will run the workflow on different occurrence datasets while listing multiple model modules will fit different models to a shared, identical dataset but then apply output modules to each model separately. In cases where modules containing randomness need to be run many times - such as running a workflow with hundreds of simulated datasets - `replicate` can be used. This simply creates a `list` of the specified module, repeated a given number of times.

The complete workflow, with multiple modules defined with the `Chain` function and parallel analyses being defined with the `list` or `replicate` function can be visualised by calling `plot` with the workflow object as it's argument.
This displays the names of all modules and the broader structure of the workflow as well as displaying which modules are not available in the online repository (Figure ).

@@

All modules, created by the zoon authors or contributed by others, have documentation explaining their methods and how they are used. This documentation is easily accessed in R by using the function `ModuleHelp`. This function fetched the documentation from the module repository.

@@

`zoon` provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform figshare. `ZoonFigshare` takes a workflow object and some minimal metadata and uploads the workflow as an RData object, along with a metadata text file, to the user's figshare profile for others to download, inspect and modify.

@@

```{r fengworkflow, eval = TRUE, dpi = 600, fig.show = "hide", results = "hide", message = FALSE, fig.height = 9, fig.width = 9}
FengPapes <- 
  workflow(occurrence = SpOcc('Dasypus novemcinctus',
                              extent = c(-130, -20, -60, 60)),
            covariate = Bioclim(extent = c(-130, -20, -60, 60),
                                layers = c(1:4, 6, 9, 10, 12, 15)),
              process = Chain(Clean, 
                              MESSMask,  
                              Background(n = 10000, bias = 200), 
                              Crossvalidate(k = 5)),
                model = MaxEnt,
               output = PrintMap(points = FALSE,
                                 threshold = 0.05))
```
